---
---
---

# Day 5 Notes

## Set up Environment

```{r setup}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

## **5 main dplyr functions**

| `filter():` subset rows in a dataset based on data values
| `arrange():` reorder the rows
| `select():` subset columns in a dataset based on column names
| `mutate():` create new variables with functions of existing variables
| `summarize():` collapse many values down to a single summary
| 
| Each of these functions work similarly:
| -    The first argument is a data frame
| -    The following arguments describe what to do with the data frame, using the variable (column) names

## Work for Today

We will be practicing with `dplyr` functions used the `Star Wars` dataset:

```{r}
starwars
head(starwars)
?starwars

```

### 1 Subset rows with `filter()`

`filter()` allows us to subset the **rows** of a dataset based on the values

```{r}
## Filter for rows with human Star Wars characters
filter(starwars,species == "Human" )

## save to a new R object
humans <- filter(starwars, species == "Human")
humans
filter(humans, sex == "male")

## filter multiple columns at once
filter(starwars, species == "Human", sex == "male")
```

We can use different comparison operators to filter:

| `!` is "not"
| `==`  is "equal
| `!=` is "not equal"
| `&` is "and" (can use a comma instead)
| `|` is "or"
| `<` is "less than"
| `>` is "greater than"
| `is.na()` returns value as true

```{r}
## return characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
## How to get all characters that are NOT human?
filter(starwars, species != "Human")
```

`variable_name %in% vector_of_values` returns `TRUE` if the value of the variable_name is equal to one of the values in vector_of_values. This is a shortcut for variable_name == vector_of_values[1] \| variable_name == vector_of_values[2]...

```{r}
skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan") 
#reutrns false because "grey" is not in vector of colors

## How would we filter for characters with brown or blond hair?
filter(starwars, hair_color %in% c("blond","brown"))
```

`grepl("character_of_interest", variable_name)` returns `TRUE` if the value of variable_name contains the characters_of_interest

```{r}
filter(starwars, grepl("blond", hair_color))

## How would we get all characters with blue in their eyes?
filter(starwars, grepl("blue", eye_color))
```

You can use `<` , `>` , `<=`, `>=` to compare numeric variables or categorical variables

| nominal categorical variables (character vectors) are ranked alphabetically
| ordinal categorical variables (factors) are ranked based on the built-in order

```{r}
## starwars characters with a name starting with letters before "D"
filter(starwars, name < "D") #ranking is based on alphabetical order

str(diamonds)
filter(diamonds, cut > "Good") #ranking is based on levels based of the ordered factor cut
```

`between(x, left, right)` is a shorthand for `x >= left & x <= right`

```{r}
between(5, 1, 7)
filter(starwars, between(mass, 50, 100))
```

#### Practice !

Find all Star Wars characters that

| Are missing a hair color
| Are shorter than 100cm and not droids
| Were born between the years 100 and 200
| Another of your choice ! :3

```{r}
#1
filter(starwars,is.na(hair_color))

#2
filter(starwars, species != "Droid" & height <= 100 )

#3
filter(starwars, between(birth_year, 100, 200))
```

### 2 Ordering Rows with `arrange()`

`arrange()` reorders rows based on the values of column(s)

```{r}
## order by name
arrange(starwars, name)
# order of characte values: aAbBcCdD

## order by species and birthyear
arrange(starwars, species, birth_year)

## use desc() to get birth years in descending order
arrange(starwars, species, desc(birth_year))
```

#### Practice !

| Arrange the starwars dataset from tallest to shortest characters
| Arrange the starwars dataset from shortest to tallest by species
| Alphabetize the Star Wars characters by their hair color then eye color 

```{r}
#1
arrange(starwars, desc(height))

#2
arrange(starwars, height, species)

#3
arrange(starwars, hair_color, eye_color)
```

### 3 Subset columns with `select()`

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```

We can also select multiple columns in a row

```{r}
#all columns between name and eye_color (inclusive)
select(starwars,name:eye_color )
```

We can also remove columns by name using a `-`

```{r}
starwars
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
```

#### Helpful Functions

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, "species", everything())

starwars_names <- sort(names(starwars))
starwars_names
select(starwars, starwars_names)
```

`starts_with("abc")` matches column names that start with "abc"

`ends_with("abc")` matches column names that end with "abc"

`contains("abc")` matches column names that contain "abc"

```{r}
select(starwars, starts_with("h"))
```

#### Practice !

| Select each character's homeworld and species. What question might this subset of variables help answer?
| Brainstorm as many ways as possible to select `hair_color` , `eye_color` , and `birth_year`

```{r}
#1
# What planets contain which species?
# Which planet has the greatest population of select species?
select(starwars, homeworld, species)

#2
select(starwars, contain("_"))
select(starwars, contains("color"), birth_year)
select(starwars, hair_color:birth_year)
select(starwars, hair_color, skin_color, eye_color, birth_year)
select(starwars, -c(name:mass, sex:starships))
```

### 4 Add New Variables with `mutate()` 

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small,
       height_m = height/100)
```

We can create multiple variables at once and we can refer to variables we just created

```{r}
mutate(starwars_small, 
       height_m = height/100,
       bmi = mass/height_m^2)

## only keep the new variables
transmute(starwars_small, 
       height_m = height/100,
       bmi = mass/height_m^2)
```

Summary statistics give one value for the entire column: `sum()`, `mean()`

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = TRUE))

mutate(starwars_small,
       row_addition = height + mass,
       summary_sum = sum(height, mass, na.rm = T))
```

Group by a variable using the `.by` parameter

```{r}
mutate(starwars_small,
       species_mass = mean(mass, na.rm = T),
       .by = species)
```

Change the type of a variable

```{r}
mutate(starwars_small,
       mass = as.character(mass))
```

#### Practice !

| Convert mass to pounds (1 kg \~ 2.2 lbs

```{r}
mutate(starwars_small,
       mass_lbs = mass*2.2)
```

### 5 Summarize data with `summarize()` 

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T))

#Group by species
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          .by = species)
```

The only columns kept are the columns you group by and the columns you create/

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = species)

#Group by multiple groups
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = c(species,sex))
```
