# Day 6 Notes

## Set Up Environment

```{r}
#install.packages("tidyr")
#install.packages("janitor")
library(tidyr)
library(janitor)
library(dplyr)
```

```{r}
?tidyr
?janitor
?dplyr
```

`clean_names(data)` Resulting names are unique and consist only of the \_ character, numbers, and letters. Capitalization preferences can be specified using the `case` parameter

| snake_case: "`snake`"
| lowerCamel: "`lower_camel`" or "`small_camel`"
| UpperCamel: "upper_camel" or "big_camel"
| ALL_CAPS: "all_caps" or "screaming_snake"
| lowerUPPER: "lower_upper"
| UPPERlower: "upper_lower"
| Sentence case: "sentence"

## Janitor

```{r}
starwars
head(starwars)
clean_names(starwars, case = "small_camel")
clean_names(starwars, case = "screaming_snake")
new_starwars <- clean_names(starwars, case = "upper_lower")
```

How would we create a new table called StarWarsWomen of just the name and species of all female Star Wars characters in order of their birth year?

```{r}
StarWarsWomen <- select(arrange(filter(starwars,sex == "female"),birth_year),name,species)
StarWarsWomen

# You can also write it like this 
StarWarsWomen <- filter(starwars, sex == "female")
StarWarsWomen <- arrange(starwars, birth_year)
StarWarsWomen <- select(starwars, name, species)
StarWarsWomen
```

### Pipe `|>`

| `function2(function(data, parameters), params2)`
| becomes:
| `data |>`
| `fuction(parameters`

```{r}
StarWarsWomen <- starwars |>
  filter(sex == "female")|>
  arrange(birth_year)|>
  select(name, species)
StarWarsWomen
```

### The `slice_` Functions

There are five handy functions that allow you to extract specific rows within each group:

| `slice_head(data, n = 1)` takes the first row from each group
| `slide_tail(data, n = 1)` takes the last row from each group
| `slice_min(data, x, n = 1)` takes the row with the smallest value of column x
| `slice_max(data, x, n = 1)` takes the row with the largest value of column x
| `slice_sample(data, n = 1)` takes one random row

You can vary n to select more than one row

or instead of `n =`, you can use `prop = 0.1` to select (e.g) 10% of the rows in each group.

**How would you use this to get a dataset of the 10 tallest Star Wars characters?**

```{r}
slice_max(starwars, height, n = 10)
# can also sort it by species or other variables, example by species:
slice_max(starwars, height, n = 10, by = species)
#if there is a tie, you can get rid of that
slice_max(starwars, height, n = 10, by = species, with_ties = F)
```

## Tidy Data

### Pivot Longer

```{r}
table4a

pivot_longer(table4a,
             cols = c(`1999`, `2000`),
             names_to = "year",
             values_to = "cases")

table4b #shows population data
pivot_longer(table4b,
             cols =  c(`1999`, `2000`),
                       names_to = "year",
                       values_to = "population")
```

### Pivot Wider

```{r}
table2 

pivot_wider(table2,
            names_from = type,
            values_from = count)
```

### Separate

```{r}
table3

separate(table3,
         rate,
         into = c("cases", "population"),
         sep = "/")
```

### Unite

```{r}
table5
tidy_table5 <- unite(table5,
      "year",
      c("century", "year"),
      sep = "")

tidy_table5 <- table5 |> 
  unite(table5,
      "year",
      c("century", "year"),
      sep = "") |>
  separate(rate,
         into = c("cases", "population"),
         sep = "/")
```

### Combining Rows

```{r}
new_data<- data.frame(country = "USA", year = "1999", cases = "1042", population = "2000000")
bind_rows(tidy_table5, new_data)
```
